---
title: "Slowdown hypothesis"
output: html_notebook
---

# Introduction 

The aim of this notebook is to extend and update the analyses presented in [White (2002)](https://onlinelibrary.wiley.com/doi/pdf/10.1111/j.1728-4457.2002.00059.x). 

The figures/analyses are as follows:

* Table 1: How well does a linear time trend in 21 high-income countries explain change in life exptectancy at birth between 1955 and 1991? 
* Figure 1: Life expectancy at birth over time, unweighted averages of 21 high-income countries, 1955-96
* Table 2: Change in age-specific death rates between 1955 and 1991, unweighted averages for 21 high-income countries
* Figure 2: Annual change in life expectancy at birth, average of 21 high-income countries, 1955-96
* Figure 3: relationship between life expectancy at birth in 1995 and annual change in life expectancy in 21 high-income countries, 1955-96
* Table 3: Life expectancy at birth and difference from average $e_{0}$ in 21 high-income countries, 1955 to 1995
* Figure 4: Variance betwene life expectancies for 21 high-income countries, 1955-96
* Figure 5: Relationship between early (1955-75) and late (1975-95) gains in life expectancy for 21 high-income countries 

All but one of the above can be replicated with the period life expectancy data alone

## Prereqs

```{r}
pacman::p_load(
  tidyverse, HMDHFDplus,
  ggrepel
)

dta_e0 <- read_rds("tidy_data/e0_period.rds")

```

## First step: define countries 

The high income countries were: 

* Australia (AUS)
* Austria (AUT)
* Belgium (BEL)
* Canada (CAN)
* Denmark (DNK)
* Finland (FIN)
* France (FRATNP)
* Germany (West) (DEUTW)
* Greece (GRC)
* Ireland (IRL)
* Italy (ITA)
* Japan (JPN)
* Netherlands (NLD)
* New Zealand (NZL_NP)
* Norway (NOR)
* Portugal (PRT)
* Spain (ESP)
* Sweden (SWE)
* Switzerland (CHE)
* United Kingdom (GBR_NP)
* United States (USA)


```{r}
high_income_countries <- 
  c(
    "AUS", "AUT", "BEL", "CAN", "DNK", "FIN",
    "FRATNP", "DEUTW", "GRC", "IRL", "ITA", "JPN",
    "NLD", "NZL_NP", "NOR", "PRT", "ESP", "SWE",
    "CHE", "GBR_NP", "USA"
  )

all_distinct_countries <- c(
  "AUS", "AUT", "BEL", "BGR", "BLR", "CAN", "CHL", "HRV", "CHE",
  "CZE", "DEUTE", "DEUTW", "DNK", "ESP", "EST", "FIN", "FRATNP", "GRC", "HUN",
  "IRL", "ISL", "ISR", "ITA", "JPN", "KOR", "LTU", "LUX", "LVA", "NLD", "NOR",
  "NZL_NP",
  "POL", "PRT", "RUS", "SVK", "SN", "SWE", "TWN", "UKR", "GBR_NP", "USA"
)

```

# Replications 

## Table 1

To do

## Figure 1

Life expectancy at birth over time, unweighted averages of 21 high-income countries, 1955-96

```{r}
fig_01_dta <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() 

fig_01_dta %>% 
  ggplot(aes(x = year, y = mean_e0)) +
  geom_point() +
  stat_smooth(method = "lm", se = F, colour = "black", linetype = "dashed") 
```

And what's the average improvement per year?

```{r}
fig_01_dta %>% 
  lm(mean_e0 ~ year, data = .) %>% 
  summary()

```

Around  0.157 years/year, not the 0.208 given in the paper. 


Now how has this continued?

```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() 

tmp %>% 
  ggplot(aes(x = year, y = mean_e0)) +
  geom_point() +
  stat_smooth(method = "lm", se = F, colour = "black", linetype = "dashed") +
  geom_vline(xintercept = 1996)
```

So the interesting aspect of this is that life expectancy trends improved in all high income countries over the 1990s and 2000s at a faster rate. Looks to be 1993 onwards 

What does this look like for the earlier and later period? 

```{r}

tmp %>% 
  lm(mean_e0 ~ year, .) %>% 
  summary()

tmp %>% 
  filter(year < 1993) %>% 
  lm(mean_e0 ~ year, .) %>% 
  summary()

tmp %>% 
  filter(year >= 1993) %>% 
  lm(mean_e0 ~ year, .) %>% 
  summary()



```

So, from 1955 to 1992, mean life expectancy increased by on average 0.159 years per year. 

But from 1993 to 2016, mean life expectancy increased by on average 0.254 years per year, with a near perfect correlation. 


Let's look at how the UK compares with this. 

```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() 

tmp %>% 
  ggplot(aes(x = year, y = mean_e0)) +
  geom_point() +
  stat_smooth(method = "lm", se = F, colour = "black", linetype = "dashed") +
  geom_vline(xintercept = 1996) +
  geom_point(
    aes(x = year, y = e0),
    inherit.aes = F, 
    data = dta_e0 %>% 
      filter(sex == "total") %>% 
      filter(code == "GBR_NP") %>% 
      filter(between(year, 1955, 2016)) ,
    colour = "red", shape = 2
  )

```

And what was the rate of improvement in UK only?

```{r}


dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(code == "GBR_NP") %>% 
  filter(between(year, 1955, 2016)) %>% 
  lm(e0 ~ year, .) %>% 
  summary()

dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(code == "GBR_NP") %>% 
  filter(between(year, 1955, 1992)) %>% 
  lm(e0 ~ year, .) %>% 
  summary()

dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(code == "GBR_NP") %>% 
  filter(between(year, 1993, 2016)) %>% 
  lm(e0 ~ year, .) %>% 
  summary()




```

### Life expectancy for best country (labelled by this country)


A similar approach to White (2002) is presented in [Christenson 2009](https://linkinghub.elsevier.com/retrieve/pii/S0140673609614604) Their approach is to look at life expectancy for the top performing country, rather than the average of these. Let's now replicate this 


```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  summarise(
    max_e0 = max(e0, na.rm = T),
    which_code = code[e0 == max(e0)]
    ) %>% 
  ungroup() 

tmp %>% 
  ggplot(aes(x = year, y = max_e0)) +
  geom_point() +
  geom_text(
    aes(x = year, y = max_e0, label = which_code), 
    inherit.aes = F,
    data = tmp %>% filter(year %in% seq(1955, 2015, by = 5)), 
    nudge_y = 0.5
    
  ) + 
  stat_smooth(method = "lm", se = F, colour = "black", linetype = "dashed") +
  geom_vline(xintercept = 1996)

```

Japan has remained the high income country with the highest life expectancy since the mid 1980s. 



What's the linear rate of improvement for the best performing line compared with the average improvement line? 

```{r}
tmp %>% 
  filter(between(year, 1955, 2016)) %>% 
  lm(max_e0 ~ year, .) %>% 
  summary()

```

So, the rate of improvement has been about 0.196 years/year for the best performing line, compared with 0.191 years/ year for the average of the 21 countries. 

Let's now plot the mean, upper and lower together. 

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  summarise(
    max_e0 = max(e0, na.rm = T),
    which_code_max = code[e0 == max(e0)],
    min_e0 = min(e0, na.rm = T),
    which_code_min = code[e0 == min(e0)],
    mean_e0 = mean(e0, na.rm = T),
    uk_e0 = e0[code == "GBR_NP"]
    ) %>% 
  ungroup() 

tmp %>% 
  ggplot(aes(x = year)) +
  geom_point(aes(y = max_e0), colour = "green") +
  geom_text(
    aes(x = year, y = max_e0, label = which_code_max), 
    inherit.aes = F,
    colour = "green",
    data = tmp %>% filter(year %in% seq(1955, 2015, by = 5)), 
    nudge_y = 0.5
  ) + 
  geom_point(aes(y = mean_e0)) +
  geom_point(aes(y = min_e0), colour = "red") +
  geom_point(aes(y = uk_e0), colour = "blue") +
  geom_text(
    aes(x = year, y = min_e0, label = which_code_min), 
    inherit.aes = F,
    colour = "red",
    data = tmp %>% filter(year %in% seq(1955, 2015, by = 5)), 
    nudge_y = - 0.5
  ) + 
  
  stat_smooth(aes(y = max_e0), method = "lm", se = F, colour = "green", linetype = "dashed") +
  stat_smooth(aes(y = min_e0), method = "lm", se = F, colour = "red", linetype = "dashed") +
  stat_smooth(aes(y = mean_e0), method = "lm", se = F, colour = "black", linetype = "dashed") +
  stat_smooth(aes(y = uk_e0), method = "lm", se = F, colour = "blue", linetype = "dashed") +

  geom_vline(xintercept = 1996)

```

In the figure above, the red dots show the worst performing of the 21 high income countries, the green dots that for the best performing country. The black dots show the average (mean) for these countries, and the blue dots show the life expectancies for the UK. Every five years, the best and worst performing country is labelled. 

```{r}
tmp %>% 
  lm(max_e0 ~ year, .) %>% 
  summary()

tmp %>% 
  lm(min_e0 ~ year, .) %>% 
  summary()

```

The average annual improvement for the best-performing country is 0.199 years/year, with an R-squared of 0.979.

For the worst-performing country, the average annual improvement is 0.300 years/year, indicating a tendency towards convergence. The R-squared is somewhat lower, at 0.966. 

The USA has been the worst performing high income country since the early 2000s, and its life expectancy has fallen steadily below the long-term trend for worst-performing countries. 

The UK's life expectancy was that projected for the worst-performing country trend in 2015, and in 2016 fell below the linear projection for the worst performing country. 

## Figure 2


```{r}
fig_02_dta <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 1996)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
fig_02_dta %>% 
  ggplot(aes(x = year, y = ch_e0)) + 
  geom_point() + geom_line()

```

And how about subsequent to this?

```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
tmp %>%
  ggplot(aes(x = year, y = ch_e0)) +
  geom_point() + geom_line() +
  geom_vline(xintercept = 1996) +
  geom_hline(yintercept = 0)

```

It seems clear that life expectancy improvements accelarted after the early 1990s in the 21 high income countries. However also that the 2014-15 fall represented the first year of falling life expectancy (on average) since 1992-3. 

Let's see how the UK compares with this 


```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
tmp2 <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  filter(code == "GBR_NP") %>% 
  mutate(ch_e0 = e0 - lag(e0))

tmp %>%
  ggplot(aes(x = year, y = ch_e0)) +
  geom_point() + geom_line() +
  geom_point(
    aes(x = year, y = ch_e0),
    colour = "red", shape = 2, 
    data = tmp2,
    inherit.aes = F
  ) + 
  geom_line(
    aes(x = year, y = ch_e0),
    colour = "red", linetype = "dashed", 
    data = tmp2,
    inherit.aes = F
  ) + 

  geom_vline(xintercept = 1996) +
  geom_hline(yintercept = 0)



```

And now what about Scotland?

```{r}


tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
tmp2 <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  filter(code == "GBR_SCO") %>% 
  mutate(ch_e0 = e0 - lag(e0))

tmp %>%
  ggplot(aes(x = year, y = ch_e0)) +
  geom_point() + geom_line() +
  geom_point(
    aes(x = year, y = ch_e0),
    colour = "blue", shape = 2, 
    data = tmp2,
    inherit.aes = F
  ) + 
  geom_line(
    aes(x = year, y = ch_e0),
    colour = "blue", linetype = "dashed", 
    data = tmp2,
    inherit.aes = F
  ) + 

  geom_vline(xintercept = 1996) +
  geom_hline(yintercept = 0)



```

How about Japan and the USA?

```{r}


tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
tmp2 <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  filter(code == "JPN") %>% 
  mutate(ch_e0 = e0 - lag(e0))

tmp3 <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  filter(code == "USA") %>% 
  mutate(ch_e0 = e0 - lag(e0))

tmp %>%
  ggplot(aes(x = year, y = ch_e0)) +
  geom_point() + geom_line() +
  geom_point(
    aes(x = year, y = ch_e0),
    colour = "blue", shape = 2, 
    data = tmp2,
    inherit.aes = F
  ) + 
  geom_line(
    aes(x = year, y = ch_e0),
    colour = "blue", linetype = "dashed", 
    data = tmp2,
    inherit.aes = F
  ) + 
  geom_point(
    aes(x = year, y = ch_e0),
    colour = "red", shape = 3, 
    data = tmp3,
    inherit.aes = F
  ) + 
  geom_line(
    aes(x = year, y = ch_e0),
    colour = "red", linetype = "dashed", 
    data = tmp3,
    inherit.aes = F
  ) + 

  geom_vline(xintercept = 1996) +
  geom_hline(yintercept = 0)



```

The departure in the form the average is very clear in this figure, with the annual changes in the USA declining after 2008, and not improving at the average rate in the early 2000s either. Since the 1980s the USA appears to slow down worse than the average in the down-turns, then not recover as quickly as the average in the up-turns! 


And the following does this for each of the countries 


```{r}
tmp <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(mean_ch_e0 = mean_e0 - lag(mean_e0)) %>% 
  select(year, mean_ch_e0)

tmp2 <- dta_e0 %>% 
  filter(sex == "total") %>%
  filter(code %in% high_income_countries) %>% 
  filter(between(year, 1954, 2016)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>%
  ungroup() %>% 
  select(year, code, ch_e0)

tmp3 <- left_join(tmp2, tmp, by = c("year" = "year"))

tmp3

tmp3 %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = mean_ch_e0), colour = "grey", alpha = 0.5) + 
  geom_line(aes(y = mean_ch_e0), colour = "grey", alpha = 0.5) + 
  geom_point(aes(y = ch_e0)) + geom_line(aes(y = ch_e0)) + 
  facet_wrap(~code) +
  geom_hline(yintercept = 0)

```


## Figure 3

```{r}
dta_for_fig03 <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_1955 = ifelse(1955 %in% year, e0[year == 1955], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_1955, mean_ch_e0) 

dta_for_fig03 %>% 
  ggplot(aes(x = val_in_1955, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 1955 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code))



```

What about for all HMD countries?



```{r}
dta_e0 %>% 
  filter(code %in% all_distinct_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_1955 = ifelse(1955 %in% year, e0[year == 1955], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_1955, mean_ch_e0) %>% 
  ggplot(aes(x = val_in_1955, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 1955 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code)) 
```

And what about more recently?

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_1955 = ifelse(1955 %in% year, e0[year == 1955], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_1955, mean_ch_e0) 

tmp %>% 
  ggplot(aes(x = val_in_1955, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 1955 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code))

```

How about since 2010 only?

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 2010, 2016)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_2010 = ifelse(2010 %in% year, e0[year == 2010], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_2010, mean_ch_e0) 

tmp %>% 
  ggplot(aes(x = val_in_2010, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 2010 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code)) +
  stat_smooth(method = "lm", se = F, colour = "black")

```

This is an important 'dog that didn't bark': there's no apparent association between the rate of slowdown since 2010 and the life expectancy in 2010. Though the long-term tendency may be for there to be a slowdown, it's not something that explains what happened this time. 

Let's see if this were the case for other five year periods

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  mutate(decade = cut(year, seq(1955, 2015, by = 10), include.lowest = T)) %>% 
  group_by(code, decade) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year))

tmp2 <- tmp %>% filter(code == "GBR_NP")

tmp %>% 
  ggplot(
    aes(x = e0, y = mean_ch_e0, group = decade, colour = decade, shape = decade)
    ) + 
  geom_point() + 
  geom_point(
    aes(x = e0, y = mean_ch_e0, group = decade, colour = decade, shape = decade),
    inherit.aes = FALSE, 
    data = tmp2,
    shape = 19, size = 5, alpha = 0.25
  ) +
  stat_smooth(method = "lm", se = F)

```
This figure shows that the degree to which the limits-to-growth argument holds in high income countries is falling, rather than increasing, again suggesting it is not an important factor in explaining the recent slowdown in the UK (highlighted with a translucent symbol). 


## Tadpole charts and derivatives 

To start let's just show e0 against change in e0 for all countries, just as points 


```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>%
  ungroup() %>% 
  ggplot(aes(x = e0, y = ch_e0)) + 
  geom_point() + 
  geom_hline(yintercept = 0) +
  stat_smooth(method = "lm", se = F)
```

Here there seems to be no relationship whatsoever. 

But what if there's a Simpson's Paradox style issue here? What do the relationships look like for each of the groups?


```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>%
  ungroup()

tmp %>% 
  ggplot(aes(x = e0, y = ch_e0, group = code)) + 
  geom_hline(yintercept = 0) +
  geom_line(
    stat = "smooth", method = "lm", se = F, 
    colour = "black", alpha = 0.5) + 
  stat_smooth(
    method = "lm", se = F, 
    data = tmp %>% filter(code == "GBR_NP")
  )

```

Once again, there's no obvious trend indicating a limit to life expectancy, and the UK is one of a number of countries where the long-term correlation between life expectancy and change in life expectancy is positive rather than negative. 

Perhaps this issue is confounded by the negative autocorrelation found between annual lags. To investigate this we could try the average of two years rather than a single year

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>%
  mutate(avg_e0 = (e0 + lag(e0) / 2)) %>% 
  mutate(lag_ch_e0 = lag(ch_e0)) %>% 
  mutate(avg_ch_e0 = (ch_e0 + lag_ch_e0) / 2) %>% 
  ungroup() %>% 
  ggplot(aes(x = avg_e0, y = avg_ch_e0)) + 
  geom_point() + 
  geom_hline(yintercept = 0) +
  stat_smooth(method = "lm", se = F)


```
Again this makes no appreciable difference

```{r}

tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>%
  mutate(lag_ch_e0 = lag(ch_e0)) %>% 
  mutate(avg_e0 = (e0 + lag(e0) / 2)) %>% 
  mutate(avg_ch_e0 = (ch_e0 + lag_ch_e0) / 2) %>% 
  ungroup()

tmp %>% 
  ggplot(aes(x = avg_e0, y = avg_ch_e0, group = code)) + 
  geom_hline(yintercept = 0) +
  geom_line(
    stat = "smooth", method = "lm", se = F, 
    colour = "black", alpha = 0.5) + 
  stat_smooth(
    method = "lm", se = F, 
    data = tmp %>% filter(code == "GBR_NP")
  )

```

Again, this does not seem to make any difference. 



## Table 3

To do or not bother. I think we have what we need for a short paper. 

## Figure 4

Variance tween life expectancies for 21 high-income countries, 1955-96

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 12))

```

First question is what does this now look like?

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 12))

```

So, there's been a steady increase in the variance between countries after 2010. *This is an important finding*. 

Secondly, what does this look like for all countries?

```{r}
dta_e0 %>% 
  filter(code %in% all_distinct_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 20))

```

This looks to be undermined by changing population composition. 


For the higher income countries in the original selection, let's plot the difference for each individual country

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup() %>% 
  ggplot(aes(x = year, y = diff_from_mean)) + 
  facet_wrap(~code) + 
  geom_line() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1996)

```

Now let's present this all at once

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup()

tmp %>% 
  ggplot(aes(x = year, y = diff_from_mean, group = code)) + 
  geom_line(alpha = 0.6) +
  geom_line(aes(x = year, y = diff_from_mean, group = code, colour = code), 
            inherit.aes = F, size = 1.5, 
            data = tmp %>% 
              filter(code %in% c("USA", "GBR_NP"))
            ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1996)


```

Both the UK and USA are distinct, but in different ways. 

The USA had been declining in relative terms for many years, with a post 2010 acceleration.

Whereas the UK had been stagnating in relative terms for many years, then falling more rapidly post 2012ish. 

Let's just replicate the above for only 1997 onwards

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1997, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup()

tmp %>% 
  ggplot(aes(x = year, y = diff_from_mean, group = code)) + 
  geom_line(alpha = 0.6) +
  geom_line(aes(x = year, y = diff_from_mean, group = code, colour = code), 
            inherit.aes = F, size = 1.5, 
            data = tmp %>% 
              filter(code %in% c("USA", "GBR_NP"))
            ) +
  geom_hline(yintercept = 0) 


```
This also seems to be an important paper, as it shows how both the UK and the USA have been falling from the rich country mean more rapidly after 2012, though faster and from a lower starting point (of longer term relative decline) for the USA. 


## Figure 5



```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
