---
title: "Slowdown hypothesis"
output: html_notebook
---

# Introduction 

The aim of this notebook is to extend and update the analyses presented in [White (2002)](https://onlinelibrary.wiley.com/doi/pdf/10.1111/j.1728-4457.2002.00059.x). 

The figures/analyses are as follows:

* Table 1: How well does a linear time trend in 21 high-income countries explain change in life exptectancy at birth between 1955 and 1991? 
* Figure 1: Life expectancy at birth over time, unweighted averages of 21 high-income countries, 1955-96
* Table 2: Change in age-specific death rates between 1955 and 1991, unweighted averages for 21 high-income countries
* Figure 2: Annual change in life expectancy at birth, average of 21 high-income countries, 1955-96
* Figure 3: relationship between life expectancy at birth in 1995 and annual change in life expectancy in 21 high-income countries, 1955-96
* Table 3: Life expectancy at birth and difference from average $e_{0}$ in 21 high-income countries, 1955 to 1995
* Figure 4: Variance betwene life expectancies for 21 high-income countries, 1955-96
* Figure 5: Relationship between early (1955-75) and late (1975-95) gains in life expectancy for 21 high-income countries 

All but one of the above can be replicated with the period life expectancy data alone

## Prereqs

```{r}
pacman::p_load(
  tidyverse, HMDHFDplus,
  ggrepel
)

dta_e0 <- read_rds("tidy_data/e0_period.rds")

```

## First step: define countries 

The high income countries were: 

* Australia (AUS)
* Austria (AUT)
* Belgium (BEL)
* Canada (CAN)
* Denmark (DNK)
* Finland (FIN)
* France (FRATNP)
* Germany (West) (DEUTW)
* Greece (GRC)
* Ireland (IRL)
* Italy (ITA)
* Japan (JPN)
* Netherlands (NLD)
* New Zealand (NZL_NP)
* Norway (NOR)
* Portugal (PRT)
* Spain (ESP)
* Sweden (SWE)
* Switzerland (CHE)
* United Kingdom (GBR_NP)
* United States (USA)


```{r}
high_income_countries <- 
  c(
    "AUS", "AUT", "BEL", "CAN", "DNK", "FIN",
    "FRATNP", "DEUTW", "GRC", "IRL", "ITA", "JPN",
    "NLD", "NZL_NP", "NOR", "PRT", "ESP", "SWE",
    "CHE", "GBR_NP", "USA"
  )

all_distinct_countries <- c(
  "AUS", "AUT", "BEL", "BGR", "BLR", "CAN", "CHL", "HRV", "CHE",
  "CZE", "DEUTE", "DEUTW", "DNK", "ESP", "EST", "FIN", "FRATNP", "GRC", "HUN",
  "IRL", "ISL", "ISR", "ITA", "JPN", "KOR", "LTU", "LUX", "LVA", "NLD", "NOR",
  "NZL_NP",
  "POL", "PRT", "RUS", "SVK", "SN", "SWE", "TWN", "UKR", "GBR_NP", "USA"
)

```

# Replications 

## Table 1

To do

## Figure 1

Life expectancy at birth over time, unweighted averages of 21 high-income countries, 1955-96

```{r}
fig_01_dta <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() 

fig_01_dta %>% 
  ggplot(aes(x = year, y = mean_e0)) +
  geom_point() +
  stat_smooth(method = "lm", se = F, colour = "black", linetype = "dashed") 
```

And what's the average improvement per year?

```{r}
fig_01_dta %>% 
  lm(mean_e0 ~ year, data = .) %>% 
  summary()

```

Around  0.157 years/year, not the 0.208 given in the paper. 

And now for all available countries 



## Figure 2


```{r}
fig_02_dta <- dta_e0 %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1954, 1996)) %>% 
  group_by(year) %>% 
  summarise(mean_e0 = mean(e0, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(year) %>% 
  mutate(ch_e0 = mean_e0 - lag(mean_e0))
  
fig_02_dta %>% 
  ggplot(aes(x = year, y = ch_e0)) + 
  geom_point() + geom_line()

```




## Figure 3

```{r}
dta_for_fig03 <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_1955 = ifelse(1955 %in% year, e0[year == 1955], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_1955, mean_ch_e0) 

dta_for_fig03 %>% 
  ggplot(aes(x = val_in_1955, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 1955 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code))



```

What about for all HMD countries?



```{r}
dta_e0 %>% 
  filter(code %in% all_distinct_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(code) %>% 
  arrange(year) %>% 
  mutate(val_in_1955 = ifelse(1955 %in% year, e0[year == 1955], NA)) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  mutate(mean_ch_e0 = mean(ch_e0, na.rm = T)) %>% 
  filter(year == min(year)) %>% 
  ungroup() %>% 
  select(code, val_in_1955, mean_ch_e0) %>% 
  ggplot(aes(x = val_in_1955, y = mean_ch_e0)) +
  geom_point() + 
  labs(x = "Life expectancy in 1955 (years)", y = "Average annual change in life expectancy") + 
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggrepel::geom_text_repel(aes(label = code)) 
```



## Table 3

## Figure 4

Variance ebtween life expectancies for 21 high-income countries, 1955-96

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 1996)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 12))

```

First question is what does this now look like?

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 12))

```

So, there's been a steady increase in the variance between countries after 2010. *This is an important finding*. 

Secondly, what does this look like for all countries?

```{r}
dta_e0 %>% 
  filter(code %in% all_distinct_countries) %>% 
  mutate(group = ifelse(!(code %in% c("GRC", "IRL", "PRT", "ESP")), "inc", "exc")) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2015)) %>% 
  group_by(group, year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>% 
  summarise(var = var(diff_from_mean)) %>% 
  ggplot(aes(x = year, y = var, group = group)) + 
  geom_point() + geom_line() +
  scale_y_continuous(limits = c(0, 20))

```

This looks to be undermined by changing population composition. 


For the higher income countries in the original selection, let's plot the difference for each individual country

```{r}
dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup() %>% 
  ggplot(aes(x = year, y = diff_from_mean)) + 
  facet_wrap(~code) + 
  geom_line() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1996)

```

Now let's present this all at once

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1955, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup()

tmp %>% 
  ggplot(aes(x = year, y = diff_from_mean, group = code)) + 
  geom_line(alpha = 0.6) +
  geom_line(aes(x = year, y = diff_from_mean, group = code, colour = code), 
            inherit.aes = F, size = 1.5, 
            data = tmp %>% 
              filter(code %in% c("USA", "GBR_NP"))
            ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1996)


```

Both the UK and USA are distinct, but in different ways. 

The USA had been declining in relative terms for many years, with a post 2010 acceleration.

Whereas the UK had been stagnating in relative terms for many years, then falling more rapidly post 2012ish. 

Let's just replicate the above for only 1997 onwards

```{r}
tmp <- dta_e0 %>% 
  filter(code %in% high_income_countries) %>% 
  filter(sex == "total") %>% 
  filter(between(year, 1997, 2016)) %>% 
  group_by(year) %>% 
  mutate(mean_e0 = mean(e0, na.rm = T)) %>% 
  mutate(diff_from_mean = e0 - mean_e0) %>%
  ungroup()

tmp %>% 
  ggplot(aes(x = year, y = diff_from_mean, group = code)) + 
  geom_line(alpha = 0.6) +
  geom_line(aes(x = year, y = diff_from_mean, group = code, colour = code), 
            inherit.aes = F, size = 1.5, 
            data = tmp %>% 
              filter(code %in% c("USA", "GBR_NP"))
            ) +
  geom_hline(yintercept = 0) 


```

## Figure 5



```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
